---
title: "Introduction + Extending the fastverse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction + Extending the fastverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(repos = c(CRAN = "https://cran.rstudio.com/"))
```

The *fastverse* is an extensible suite of R packages, developed by independent people,
that contribute towards the objectives of:

- Speeding up R through heavy use of compiled code (C, C++, Fortran)
- Enabling more complex statistical and data manipulation operations in R
- Reducing the number of dependencies required to do advanced computations in R

The *fastverse* installs 6 core packages (*data.table*, *collapse*, *matrixStats*, *kit*, *magrittr* and *fst*) that are (by default) loaded and harmonized with `library(fastverse)`. These packages were selected because they provide compiled code for most common statistical and data manipulation tasks, have carefully managed API's, jointly depend only on base R and *Rcpp*, and work remarkably well together. 

```{r}
library(fastverse)
```

The package then provides functionality familiar from the *tidyverse* package, such as checking and reporting namespace clashes, and utilities to update the packages. 

```{r}
# Checking for any updates
fastverse_update()
```

While placing a clear emphasis on High-Quality, High-Performance and Low-Dependency code, which warranted a selection of core packages and a number of [suggested extensions](link), the *fastverse* is not particularly opinionated and can be freely and permanently extended by any number of packages. These extensions can then be loaded and managed by the *fastverse* package, just like the core packages. 

## Extending the *fastverse* for the Session
After the core packages have been loaded with `library(fastverse)`, it is possible to extend the *fastverse* for the current session by adding any number of additional packages with `fastverse_extend()`. This will attach the packages and (default) check for function clashes with core packages and among the extension packages. 

```{r}
# Extend the fastverse for the session
fastverse_extend(xts, roll, fasttime)

# See that these are now part of the fastverse
fastverse_packages()

# They are also saved in a like-named option 
options("fastverse_extend")

# Check versions and update status of packages and dependencies
fastverse_sitrep()
```

We can detach all fastverse packages (or particular packages) using `fastverse_detach`.

```{r}
# Detaches all packages (including itself) but does not (default) unload them
fastverse_detach()
```

Note that for programming purposes it is also possible to pass vectors of packages to both `fastverse_extend` and `fastverse_detach`^[In particular, the `...` expression is first captured using `substitute(c(...))`, and then evaluated inside `tryCatch`. If this evaluation fails or did not result in a character vector, the expression is coerced to character.]. The defaults of `fastverse_detach` are set such that detaching is very 'light'. Packages are not unloaded and all *fastverse* options set for the session are kept.

```{r}
# Our extensions are still here
options("fastverse_extend")

# Thus attaching the fastverse again will include them
library(fastverse)
```

'Harder' modes of detaching can be achieved using arguments `unload = TRUE` (and `force = TRUE`) to (forcefully) detach and unload *fastverse* packages, and/or `session = TRUE` which will clear all *fastverse* options set^[`options("fastverse_quiet")` will only be cleared if all packages are detached]. 
```{r}
# Detaching and unloading all packages and clearing opions
fastverse_detach(session = TRUE, unload = TRUE)
```

`fastverse_detach` can also be used to detach any other packages that are not part of the *fastverse*. 

Since `options("fastverse_extend")` keeps track of which packages were added to the *fastverse* for the current session, it is also possible to set it before loading the fastverse e.g. 

```{r}
options(fastverse_extend = c("ggplot2", "scales"))
library(fastverse)

fastverse_detach(session = TRUE)
```

## Permanent *fastverse* Modifications

`fasvtverse_extend` and `fastverse_detach` both have an argument `permanent = TRUE` which can be used to make the changes persist across R sessions. This is implemented using a file called 'fastverse_extend.txt' which is saved in the package directory (so it will be removed when the *fastverse* is reinstalled). 

For example, suppose most of my work involves time series analysis, and I would like to add *xts*, *roll*, *forecast*, *tsbox* and *dygraphs* to my *fastverse*. Let's say I also don't really use the *fst* file format, and I don't really need *matrixStats* either as I can do most of the statistics I need with base R and *collapse*. Let's finally say that I don't want `xts::first` and `xts::last` to mask `data.table::first` and `data.table::last`.

Then I could permanently modify my *fastverse* as follows: 

```{r}
library(fastverse)
# Adding extensions
fastverse_extend(xts, roll, forecast, tsbox, dygraphs, permanent = TRUE)
# Removing some core packages
fastverse_detach(data.table, fst, matrixStats, permanent = TRUE)
# Adding data.table again, so it is loaded last
fastverse_extend(data.table, permanent = TRUE)
```

To verify our modification, we can see the order in which the packages are loaded, and do a conflict check:

```{r}
# This will be the order in which packages are loaded
fastverse_packages(include.self = FALSE)
# Check conflicts to make sure data.table functions take precedence
fastverse_conflicts()
```

I also note here that `options("fastverse_extend")` is still empty, because we have written those changes to a file. After this permanent modification I can still extend the *fastverse* for the session:

```{r}
# Extension for the session
fastverse_extend(ggplot2, scales)
# These packages go here
options("fastverse_extend")
# This fetches packages from both the file and the option
fastverse_packages()
```

Now lets see if our permanent modification worked: 

```{r}
# detach all packages and clear all options
fastverse_detach(session = TRUE)
library(fastverse) 
```

As long as the current installation of the *fastverse* is kept, these modifications will persist across R sessions. 

## Custom *fastverse* Configurations for Projects

An additional bit of flexibility afforded by the *fastverse* is the ability to put together a custom collection of packages for a project that you want to load and manage with `library(fastverse)`.

You can do this by including a configuration file named `.fastverse` (no file extension) inside a project directory, and placing inside that file the names of packages to be loaded when calling `library(fastverse)`^[You can place package names in that file any manner you deem suitable: separated using spaces or commas, on one or multiple lines. Note that the file will be read from left to right and from top to bottom. Packages are loaded in the order found in the file.]. This will ignore any global configuration discussed in the previous section. You can still extend the *fastverse* inside a project session using `fastverse_extend` or `options("fastvers_extend")`. 

## Quiet Mode

A useful option that was not yet demonstrated is `option("fastverse_quiet")`:

```{r}
fastverse_detach()
options(fastverse_quiet = TRUE)
library(fastverse)

# This gives lots of function clashes with data.table, but they are not displayed in quiet mode
fastverse_extend(lubridate)
```

If we only want to omit a function clash check when calling `fastverse_extend`, we can also use `fastverse_extend(..., check.conflicts = FALSE)`. I note at this point it is not possible to permanently set `options("fastverse_quiet")`. 

## Conclusion

The *fastverse* package exists principally for 2 reasons: to promote quality high-performance software for R, and to provide a flexible approach to package loading and management in R. This vignette introduced the key features of the package. Suggested *fastverse* extension packages are introduced in the [README](yyy) file. 

```{r}
# Resetting the fastverse to defaults (clearing all permanent extensions and options)
fastverse_reset()
# Detaching 
fastverse_detach()
```





